# GapMinder Computations

## Load & Check Data

```{r gapMinder, echo = TRUE}
library(tidyverse)

gapMinderSet <- read_delim("gapminder.csv")


ncol(gapMinderSet)
nrow(gapMinderSet)
dim(gapMinderSet)

gapMinderSet %>% 
  head(3)
```

1.  There is a total of **13,055** rows of data observations, and **25** column vectors, each possessing its own unique field name.

2.  After providing the first 3 lines of the gap minder set in the code snippet above, the data frame follows the visual aspect of a data frame,      but is **contains** a vast amount of **NA** values for most of the vector columns.


## Descriptive Statistics

```{r}

gapMinderSet <- read_delim("gapminder.csv")

gapMinderSet %>% 
  select(iso2, iso3, name) %>% 
  summarize(totalCountries = n_distinct(name), totalISO2 = n_distinct(iso2),
            totalISO3 = n_distinct(iso3))

```

**Part**: 2A
```{r}

gapMinderSet <- read_delim("gapminder.csv")

gapMinderSet %>% 
  select(iso2, name) %>%
  group_by(iso2) %>% 
  summarize(namesISO2 = n_distinct(name)) %>% 
  arrange(desc(namesISO2))

gapMinderSet %>% 
  select(iso2, name) %>% 
  filter(is.na(iso2)) %>% 
  distinct(name)

```
**Explanation**: The iso-2 codes that corresponds to more than one name is: **Namibia**.


**Part**: 2B
```{r}

gapMinderSet <- read_delim("gapminder.csv")

gapMinderSet %>% 
  select(iso3, name) %>%
  group_by(name) %>% 
  summarize(namesISO3 = n_distinct(iso3)) %>% 
  arrange(desc(namesISO3))

gapMinderSet %>% 
  filter(is.na(name)) %>% 
  distinct(iso3)
```

**Explanation**: These list of countries include: **CHANSIL**, **GBM**, **KOS**, **NLD_CURACAO**


```{r}
gapMinderSet %>% 
  summarize(max = max(time, na.rm = TRUE), min = min(time, na.rm = TRUE))
```


## CO2 Emissions

#### Question 1:
```{r}
gapMinderSet %>% 
  select(time, co2, co2_PC) %>% 
  group_by(time) %>% 
  summarize(co2Missing = sum(is.na(co2)), co2_PCMissing = sum(is.na(co2_PC)))
  
```

#### Question 2:
```{r}

gapMinderSet %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | iso3 == "IND" | 
         iso3 == "RUS" | iso3 == "JPN") %>% 

ggplot() +
  geom_line(aes(time, co2, col = name)) +
  labs(x = "Year",
       y = "CO2 Emissions")

```

**Explanation**: After looking at the line graph, it's a fair argument that CO2 emissions have **increased** for the selected countries as time progressed. However, the strength of the increasing pattern *differs* for each of the selected countries. For instance, China has had the strongest *sudden* rise in CO2 emissions from the year **2000** to the approximately **2019**. On the other hand, Japan has had the weakest flow of increase in CO2 emissions over the course of time.

#### Question 3: 
```{r}
gapMinderSet %>% 
  filter(iso3 == "USA" | iso3 == "CHN" | iso3 == "IND" | 
         iso3 == "RUS" | iso3 == "JPN") %>% 

ggplot() +
  geom_line(aes(time, co2_PC, col = name)) +
  labs(x = "Year",
       y = "CO2 Emissions Per Capita")
```

**Explanation**: This graph is suggesting that the **United States** and the **Russian Federation** have the highest CO2 emissions per capita, despite the *minor decrease* for the United States and the *drastic decrease* of the Russian Federation. On the other hand, the two lowest countries with the lowest CO2 emissions per capita is China and India.

#### Question 4: 
```{r}
dataSet <- gapMinderSet %>% 
  select(region, co2_PC, time) %>% 
  group_by(region, time) %>% 
  filter(!is.na(region), time %in% c(1960, 2016), !is.na(co2_PC)) %>% 
  summarize(average = mean(co2_PC, na.rm = TRUE))
dataSet
```

**Explanation**: Aside from **American** region, all of the continents/regions displayed an **increase** in average CO2 emissions per capita throughout the course of time. Providing emphasis on the **American** region, the average of its CO2 emissions per capita has decreased, breaking the common pattern of the other continents.

#### Question 5:
```{r}
ggplot(dataSet, aes(region, average, fill = factor(time))) +
  geom_col(pos = "dodge") +
  labs(x = "Continent",
       y = "Average CO2 Per Capita",
       fill = "Year")
```

#### Question 6:
```{r}
highestEmitter <- gapMinderSet %>% 
  filter(time == 2016, !is.na(co2_PC), !is.na(region)) %>% 
  select(region, name, co2_PC) %>% 
  group_by(region) %>% 
  filter(rank(desc(co2_PC)) < 4) %>% 
  arrange(region, desc(co2_PC))
highestEmitter

lowestEmitter <- gapMinderSet %>%
  filter(time == 2016, !is.na(co2_PC), !is.na(region)) %>% 
  select(region, name, co2_PC) %>% 
  group_by(region) %>% 
  filter(rank(co2_PC) < 4) %>% 
  arrange(region, desc(co2_PC))
lowestEmitter
```


## GDP Per Capita

#### Question 1:
```{r}
gapMinderSet %>% 
  filter(time == 1960) %>% 
  ggplot(aes(GDP_PC, lifeExpectancy, color = region, size = totalPopulation)) +
  geom_point() +
      labs(x = "GDP Per Capita",
         y = "Life Expectancy")

```

**Explanation**: One of the first components of the graph that captured my attention was that **Asia** has had a few samples of its *dots* be the **largest** compared to its surrounding data points. Another component of the graph that captures my attention is the cluster of *dots* of the **African** region in the beginning (close to the origin).

#### Question 2:
```{r}
gapMinderSet %>% 
  filter(time == 2019) %>% 
  ggplot(aes(GDP_PC, lifeExpectancy, color = region, size = totalPopulation)) +
  geom_point() +
    labs(x = "GDP Per Capita",
         y = "Life Expectancy")
```

#### Question 3:

**Explanation**: After analyzing the two plots, the 2019 graph seems to have more **clusters** of data in comparison to the 1960 graph. One of the important development difference I noticed from 1960 to 2019 is that the **domain** of GDP Per Capita has increased tremendously.

#### Question 4: 
```{r}
gapMinderSet %>% 
  filter(time %in% c(1960, 2019)) %>% 
  group_by(region, time) %>% 
  summarize(averageLifeExpectancy = mean(lifeExpectancy, na.rm = TRUE))
```

#### Question 5: 
```{r}
lifeExpectancySet <- gapMinderSet %>% 
  filter(time %in% c(1960, 2019)) %>% 
  group_by(region, time) %>% 
  summarize(averageLifeExpectancy = mean(lifeExpectancy, na.rm = TRUE)) %>% 
  group_by(region) %>% 
  summarize(difference = averageLifeExpectancy - lag(averageLifeExpectancy)) %>% 
  filter(!is.na(difference)) %>% 
  arrange(difference)
lifeExpectancySet
```

#### Question 6:
```{r}
gapMinderSet %>% 
  filter(time == 1960 | time == 2019, !is.na(GDP_PC)) %>% 
  ggplot(aes(GDP_PC, fill = factor(time))) +
  geom_histogram(pos = "dodge") +
  labs(x = "GDP Per Capita",
       y = "Count",
       fill = "Time")

```

#### Question 7:
```{r}
gapMinderSet %>%
  filter(!is.na(name), !is.na(lifeExpectancy)) %>% 
  group_by(time) %>% 
  mutate(lifeExpectancyRank = rank(desc(lifeExpectancy))) %>% 
  ungroup() %>% 
  filter(time == 1960 | time == 2019, iso3 == "USA") %>% 
  select(name, time, lifeExpectancyRank)
  
```

#### Question 8:
```{r}
gapMinderSet %>%
  filter(!is.na(name), !is.na(lifeExpectancy)) %>% 
  group_by(time) %>% 
  mutate(lifeExpectancyRank = rank(desc(lifeExpectancy)),
         totalCountries = n(),
         LeRankPercentage = lifeExpectancyRank / totalCountries * 100) %>% 
  ungroup() %>% 
  filter(time == 1960 | time == 2019, iso3 == "USA") %>% 
  select(name, time, lifeExpectancyRank, LeRankPercentage)
```

**NOTE**: For the "LeRankPercentage" variable in the code above, I chose to multiply the percentage by 100 in order to get a returned numerical value that is between 0-100.

**Final Message**: I spent a total of 15 hours on this assignment. The textbook was supplemental when it comes to running into errors and re-framing the structure of the code scripts to answer the questions properly to the best of my capabilities.
